---
title: "ANOVA"
---

### Objective

Students perform an ANOVA analysis and interpret the results.

## What's in a name?

[ANOVA](http://www.sthda.com/english/wiki/one-way-anova-test-in-r) stands for [**An**]{.underline}alysis [**o**]{.underline}f [**Va**]{.underline}riance. It's very much like the *t*-test and Wilcoxon test, but with three or more levels (aka. groups) in the categorical variable.

|                                   |                                                           |
|-----------------------------------|-----------------------------------------------------------|
| ***t*****-test or Wilcoxon test** | **ANOVA**                                                 |
| old v. young                      | old v. middle age v. young                                |
| women v. men                      | women v. men. v. kids                                     |
| Mandarin v. Cantonese             | Mandarin v. Cantonese v. Wu v. Hokkien v. Gan v. Hunanese |

An ANOVA calculates an *F* statistic which, at a high level is calculated with the following formula:

$$
\frac{variation\ between\ groups}{variation\ within\ groups}
$$

The hypotheses are:

$$
H_0:μ_1=μ_2=μ_3...=μ_n
$$

$$
H_a:At\ least\ one\ of\ the\ groups\ is\ different\ from\ the\ others.
$$

```{r}
library("tidyverse")
library("joeysvowels")
front_vowels <- joeysvowels::midpoints %>%
  filter(vowel %in% c("TRAP", "FACE", "DRESS", "FLEECE", "KIT")) %>% 
  mutate(dur = end - start)
front_vowels %>% 
  ggplot(aes(x = vowel, y = dur))+
  geom_boxplot(notch = TRUE)
```

Those notches are trying to tell a story, amirite?!

Let's see what an ANOVA says:

```{r}
front_vowels %>% 
  aov(dur~vowel, data = .) -> result
summary(result)
```

Okay, the ANOVA feels that there is something goin' on here. Just look at that *p*-value well below 0.05. But, the immediate follow-up question is which groups are significantly different from other groups.

Enter the post-hoc test. Actually, there are quite a few post-hoc test, but we'll use the Tukey Honest Significant Difference post-hoc test (because that's what Dr. Brown sees a lot in publications):

```{r}
TukeyHSD(result)
```

Simple enough. The *p*-value tell of which pairwise comparisons are statistically significantly different from each other.

## Activity

Your turn! Find a dataset of your choice and load it up, and find a continuous variable and a categorical variable with at least three levels, and ... you guessed, run an ANOVA.
