---
title: "mixed-effects-logistic"
---

## Objective

Students fit a mixed-effects logistic regression model.

## Logistic regression with random effects

As we saw in a [previous lesson](https://ekbrown.github.io/ling_data_analysis/lessons/mixed-effects-linear.html), random effects control for the variability in the response variable that comes from the individual members of a population that we happen to get into our sample. The `lme4` R package has a [`glmer()`](https://www.rdocumentation.org/packages/lme4/versions/1.1-35.1/topics/glmer) function that can be used to fit a mixed-effects logistic regression when you specify \`family = "binomial"\`.

Let's take a look at a dataset used to write [an article](https://doi.org/10.1515/cllt-2020-0016) about the effect of fast speech on word durations and the durations of the /s/ segment in a sample of Colombian Spanish. It's named "Dataset_fast_spch_Colombia.xlsx" and is in the LMS.

```{r}
suppressPackageStartupMessages(library("tidyverse"))
suppressPackageStartupMessages(library("lme4"))
library("readxl")

colombia <- read_excel("/Users/ekb5/Documents/LING_440/datasets/Dataset_fast_spch_Colombia.xlsx", sheet = "dataset")

# make the response variable a factor
colombia <- colombia %>% 
  mutate(pronunciation = as.factor(pronunciation))
```

In order to make sure we understand what the coefficients in the Estimate column mean in the regression results below, let's print out the levels of the response variable:

```{r}
levels(colombia$pronunciation)
```

So, because the second level (i.e., `maintained`) is compared to the first (i.e., `deleted`), positive coefficients in the Estimate column mean that `maintained` is favored, while negative coefficients in that column mean that `deleted` is favored.

Let's fit a mixed-effects logistic regression model!

```{r}
#| eval: true
### generalized linear regression (aka. logistic regression)
in_formula <- formula(pronunciation ~ 
            (1 | file) + (1 | wd) + # random effects (intercepts only)
            frc + # fixed effects
            spch_rate +
            wd_freq +
            pre_sound +
            wd_len +
            position)

m1 <- lme4::glmer(formula = in_formula, data = colombia, family = "binomial", control=glmerControl(optimizer="bobyqa", optCtrl=list(maxfun=2e4)))

summary(m1)
performance::r2_nakagawa(m1)
```

## Activity

Run the above code to fit a mixed-effects logistic regression.
