---
title: "Visualization"
---

## Objective

-   Students visualize data with easy-to-interpret plots.

## Background

R kicks trash with visualizing data. And the go-to package for creating professional-looking plots is [`ggplot2`](https://ggplot2.tidyverse.org/), one of the core packages of the `tidyverse` ecosystem. Base R also creates plots, but anybody who's anybody is using `ggplot2`. (Okay, maybe I'm overstating this a bit, but just you wait and see if I'm lying.)

## Basics

First, we need some data in a data frame (e.g., a tibble). Let's use the `Data_ptk.xlsx` file in the LMS:

```{r}
#| eval: false
library("tidyverse")
library("readxl")
ptk <- read_excel("../../../data_analysis/datasets/Data_ptk.xlsx", sheet = "data")
glimpse(ptk)
head(ptk)
```

The function `ggplot()` (no 2, like in the package name) creates a plotting window or palette:

```{r}
#| eval: false

ggplot()
```

The first argument to `ggplot()` is a data frame, which can be piped in:

```{r}
#| eval: false

ggplot(ptk)

# same as above, but with a pipe
ptk |> ggplot()
```

The second argument to `ggplot()` is a call to the `aes()` function (short for "aesthetics") which maps columns in the data frame to elements on the plot. The first argument to `aes()` is what will be plotted on the x-axis and the second argument, if needed, is what will be plotted on the y-axis. For example, let's put the levels of `LANG` on the x-axis:

```{r}
#| eval: false

ptk |> 
  ggplot(aes(x = LANG))
```

Or we can put the `prePhonBin` on the y-axis:

```{r}
#| eval: false

ptk |> 
  ggplot(aes(y = prePhonBin))
```

You'll notice that the above plots don't actually plot anything yet; they simply set up the plotting window. We now need to add elements to the plot.

## One categorical variable

Perhaps the most basic plot is of one categorical variable. In order to see the distribution of such a variable across its levels (aka. values or groups), let's create a popular option: the good ol' fashioned barplot. Note: The several layers of the plot are separated by a plus sign `+` rather than the pipe operator, and the layers are often specified with a `geom_...()` function (short for geometric object). See the section about [layers](https://ggplot2.tidyverse.org/reference/#layers) in the reference manual of ggplot2.

```{r}
#| eval: false
ptk |> 
  ggplot(aes(x = LANG))+
  geom_bar()
```

It might be good to also get the exact number of tokens in each level with our ol' friends:

```{r}
#| eval: false
ptk |> 
  count(LANG)
```

### Activity

-   Take a look at the variety of barplots that are available in [`ggplot`](https://ggplot2.tidyverse.org/reference/geom_bar.html).

-   Then, using the `Data_ptk.xlsx` dataset, plot the distribution of several categorical variables, for example: `LANG`, `GENRE`, `prePhonBin`, `folPhonBin`, `prevMention`, `cogStatus`, `lexClass`, `wdClass`.

## Two categorical variables

The following are popular plots to visualize the distribution of data points across two categorical variables.

### Stacked barplot

To create a stacked (aka. filled) barplot, one categorical variable is given as `x` in `aes()` and the other is given as the fill argument in `aes()`, for example:

```{r}
#| eval: false

ptk |> 
  ggplot(aes(x = LANG, fill=prePhonBin))+
  geom_bar()
```

See more examples of barplot in the [docs](https://ggplot2.tidyverse.org/reference/geom_bar.html).

Follow-up question: How can we get the exact numbers of tokens in each of the four subgroups displayed in the plot? After a good-faith effort, if you need help, take a look at Dr. Brown's code below:

```{r}
#| eval: false
#| code-fold: true

ptk |> 
  count(LANG, prePhonBin)
```

### Percent barplot

To create a percent barplot, that is, a stacked barplot that sums to 1.0, let's modify the previous barplot by adding the argument `position = "fill"` to `geom_bar(), but`outside of `aes()` (which isn't used at all).

```{r}
#| eval: false

ptk |> 
  ggplot(aes(x = LANG, fill=prePhonBin))+
  geom_bar(position = "fill")
```

We notice that the y-axis is labeled "count", but those aren't counts, they're proportions. Let's change the axis labels with the `labs()` element:

```{r}
#| eval: false

ptk |> 
  ggplot(aes(x = LANG, fill=prePhonBin))+
  geom_bar(position = "fill")+
  labs(x = "Language", y = "Proportion")
```

### Grouped barplot

To create a grouped barplot, modify the previous barplot by changing the argument to `position = "dodge"`. Simple enough.

```{r}
#| eval: false

ptk |> 
  ggplot(aes(x = LANG, fill=prePhonBin))+
  geom_bar(position = "dodge")
```

### Activity

-   Give it a whirl! Using the `Data_ptk.xlsx` dataset, plot the distribution of sets of two categorical variables, e.g., `LANG` and `GENRE`, `prePhonBin` and `folPhonBin`, `prevMention` and `cogStatus`, `lexClass` and `wdClass`, and any combination of these variables.

-   If you'd like, take a look at mosaic plots in the [`ggmosaic`](https://cran.r-project.org/web/packages/ggmosaic/) R package. Mosaic plots are another way to visualize two categorical variables. Look at the vignette of `ggmosaic` for examples of how to make this type of plot in the ggplot framework.

## One continuous variable

It is useful to visualize the distribution of continuous variables in order to see whether the variable is distributed normally or has some skew or is bimodal.

The following plots are useful for this purpose. You only need to specify the `x` argument in `aes()` within `ggplot()` with the one variable.

### Histogram

A histogram slices or bins of a continuous variable into N bins and then plots the count within each bin on the y-axis. It gives a barplot (i.e., it looks like a barplot), but instead of levels within a categorical variable, the bars are equidistant intervals across a continuous variable. Here's an example with the continuous variable `VOT` in the `ptk` dataset:

```{r}
#| eval: false
ptk |> 
  ggplot(aes(x = VOT))+
  geom_histogram()
```

You can specify a specific number of bins with the `bins` argument, for example:

```{r}
#| eval: false
ptk |> 
  ggplot(aes(x = VOT))+
  geom_histogram(bins = 15)
```

### Density plot

Density plots are like histograms, but they use a curved line rather than bins:

```{r}
#| eval: false
ptk |> 
  ggplot(aes(VOT))+
  geom_density()
```

The cool thing about density plots is that we can specify that the area below the curve sum to 1. That fact will be important when we look at p-values.

Let's plot `VOT` as a density plot:

```{r}
#| eval: false
ptk |> 
  ggplot(aes(scale(VOT)))+
  geom_density(fill = "blue")
```

### Activity

Give it a go! Plot the distribution of several different continuous variables in the `ptk` dataset, and be ready to share with a neighbor and/or the class.

MORE TO COME REAL SOON!
